---
title: 'Primer of Ecology with $\textsf{R}$: chapter 4 Problems (Lumpies)'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    toc: false
    code_folding: hide
    number_sections: false
    theme: cerulean
    highlight: haddock
  pdf_document:
    toc: false
---

``` {r setup, echo = TRUE, message = FALSE, warning = FALSE}

knitr::opts_chunk$set(fig.width = 6, fig.height = 4, # fig.path = 'Figs/',
                      echo = TRUE, message = FALSE, warning = FALSE)

library(tidyverse)
library(RColorBrewer)

### The book has an R package! 
library(primer) # install.packages('primer')

```

# Chapter 4: Populations in Space{.tabset}

## 4.1. Equilibria

Derive expressions and calculate equilibria for the following metapopulation models, with $c_i = 0.05$, $e = 0.01$. Show your work — start with the differential equations, set to zero, and solve $p^{∗}$; then substitute in values for $c_i, e$.

(a) Levins model.
(b) Propagule rain model (`gotelli`).
(c) Propagule rain model that also includes both external and internal propagule production and dispersal.
(d) Hanski model.
(e) Lande (habitat destruction) model (with `D = 0.1`).

### Molly's attempts:
*currently on paper... will see if I have time to practice my LaTeX :)*

### Casey's attempts:

a) Levins:
\begin{align*}
  \frac{d p}{d t} &= c_i p (1 - p) - ep  &\text{(start w/classic Levins)}\\
  0 &= c_i p (1 - p) - ep              &\text{(set FOC to zero)}\\
  0 &= p(c_i(1 - p) - e)               &\text{(rearrange)}\\
  0 &= p(c_i - c_i p - e)              &\text{(distribute)}\\
  \Longrightarrow p^* &= \frac{c_i - e}{c_i}  \text{ (or $p^* = 0$)}
\end{align*}

b) Gotelli (propagule rain):
\begin{align*}
  \frac{d p}{d t} &= c_e (1 - p) - ep     &\text{(start w/propagule rain)}\\
  0 &= c_e (1 - p) - ep                &\text{(set FOC to zero)}\\
  0 &= c_e - p(c_e - e)                &\text{(distribute and arrange)}\\
  \Longrightarrow p^* &= \frac{c_e}{c_e - e}
\end{align*}
Note this doesn't have a trivial solution at $p = 0$ since external supply of propagules!

c) Fancy propagule rain:
\begin{align*}
  \frac{d p}{d t} &= (c_i p + c_e) (1 - p) - ep   &\text{(start w/fancy propagule rain)}\\
  0 &= (c_i p + c_e) (1 - p) - ep              &\text{(set FOC to zero)}\\
  0 &= c_i p + c_e - c_i p^2 - c_e p - ep      &\text{(distribute)}\\
  0 &= -c_i p^2 + p(c_i - c_e - e) + c_e       &\text{(arrange as quadratic)}
\end{align*}
This can be solved with the quadratic formula, setting $a = -c_i, \; b = (c_i - c_e - e), \; c = c_e$.  But I'm gonna not do that here...

d) Hanski model (core-satellite, simple rescue effect):
\begin{align*}
  \frac{d p}{d t} &= c_i p (1 - p) - ep(1 - p)   &\text{(start w/Hanski)}\\
  0 &= c_i p (1 - p) - ep(1 - p)              &\text{(set FOC to zero)}\\
  0 &= c_i p - c_i p^2 - ep + ep^2            &\text{(distribute)}\\
  0 &= p[(e - c_i) p + (c_i - e)]             &\text{(arrange)}\\
  \Longrightarrow p^* &= \frac{c_i - e}{c_i - e} = 1 \text{ (or $p^* = 0$)}
\end{align*}
Note: these solutions were shown in the text; $p^* = 1$ is stable if $c_i > e$, and $p^* = 0$ is stable for $c_i < e$.

e) Lande (habitat destruction model):

\begin{align*}
  \frac{dp}{dt} &= c_i p(1 - D - p) - ep   &\text{(start w/Lande)}\\
  0 &= c_i p(1 - D - p) - ep              &\text{(set FOC to zero)}\\
  0 &= p(c_i - c_i D - c_i p - e)         &\text{(arrange)}\\
  \Longrightarrow p^* &= 1 - D - \frac{e}{c_i} \text{ (or $p^* = 0$)}
\end{align*}

## 4.2. Habitat destruction

Compare different levels of habitat destruction.

(a) Use the habitat destruction model (`lande`) to compare 9 levels of destruction (`ds <- seq(0,.8, by=.1)`), using $c_i = 0.1, e = 0.01$. Plot of graph of the dynamics through time, and calculate the equilibria directly.
(b) Write an ODE function for a habitat destruction model with rescue effect. Let the “rescue” have an additional parameter, $a$, such that extinction rate is $ep(1 − ap)$.
(c) Let $D = 0.5, c_i = 0.1, e = 0.02$, and vary $a$ over five levels (including $a = 0, 1$) to investigate the effects of “relative rescue effect” on the equilibria and dynamics of a metapopulation.

### Molly's attempts:
(a) Use the Lande model to compare 9 levels of destruction (`ds <- seq(0,.8, by=.1)`), using $c_i = 0.1, e = 0.01$. Graph the dynamics through time.
```{r}
### copied pretty much exactly from the text...
library(deSolve)
prmsD <- c(ci = 0.1, e = 0.01, D = 0)
Ds <- seq(0,.8, by=.1)
Initial.p <- 0.01
t <- 1:200

lande <- function(t, y, parms) {
  p <- y[1]
  with(as.list(parms), {
    dp <- ci * p * (1 - D - p) - e * p
    return(list(dp))
  })
}

ps <- sapply(Ds, function(d) {
  prmsD["D"] <- d
  ode(y = Initial.p, times = t, func = lande, parms = prmsD)[, 2]
})

matplot(t, ps, type = "l", ylab = "p", xlab = "time")
text(c(200, 200, 200), ps[200, ], paste("D = ", Ds, sep = ""),
     adj = c(1, 0))
```

(b) Write an ODE function for a habitat destruction model with rescue effect. Let the “rescue” have an additional parameter, $a$, such that extinction rate is $ep(1 − ap)$.
```{r}
lande_rescue <- function(t, y, parms) {
  p <- y[1]
  with(as.list(parms), {
    dp <- ci * p * (1 - D - p) - e * p * (1 - a * p)
    return(list(dp))
  })
}
```

(c) Let $D = 0.5, c_i = 0.1, e = 0.02$, and vary $a$ over five levels (including $a = 0, 1$) to investigate the effects of “relative rescue effect” on the equilibria and dynamics of a metapopulation.
```{r}
prmsA <- c(ci = 0.1, e = 0.02, D = 0.5, a = 0)
as <- seq(0, 1, by = .25)
Initial.p <- 0.01
t <- 1:200

ps2 <- sapply(as, function(a) {
  prmsA["a"] <- a
  ode(y = Initial.p, times = t, func = lande_rescue, parms = prmsA)[, 2]
})

matplot(t, ps2, type = "l", ylab = "p", xlab = "time")
text(c(200, 200, 200), ps[200, ], paste("D = ", Ds, sep = ""),
     adj = c(1, 0))
```

### Casey: 

a) 

```{r}
### Write Lande function to accept various values of D for ODE solver
lande_cco <- function(t, y, parms) {
  p <- y[1]
  with(as.list(parms), {
    dp_dt <- c_i * p * (1 - D - p) - e * p
    return(list(dp_dt))
  })
}

### set up vector of Ds and initial vals
D_vec <- seq(0, .8, .1)
t_vec <- 1:150
p_0   <- .15

params <- c(c_i = 0.1, e = 0.01, D = 0)

### Run ODE on the whole shebang
p_list <- lapply(D_vec, function(D) {
  # D <- D_vec[2]
  params$D <- D
  p_vec <- ode(y = p_0, times = t_vec, func = lande_cco, parms = params) %>%
    as.data.frame() %>%
    setNames(c('t', 'p')) %>%
    mutate(D = D)
})

p_df <- bind_rows(p_list)

ggplot(p_df, aes(x = t, y = p, color = D)) +
  geom_line(data = p_df %>% filter(D == .5), 
            color = 'red4', alpha = .5, size = 3) +
  geom_line(aes(group = D)) +
  theme_minimal() +
  labs(x = 'Time',
       y = 'Proportion of populated sites')

```

b)
```{r}
### Write Lande function to accept various values of D for ODE solver
lande_a_cco <- function(t, y, parms) {
  p <- y[1]
  with(as.list(parms), {
    dp_dt <- c_i * p * (1 - D - p) - e * p * (1 - a * p)
    return(list(dp_dt))
  })
}
```

c)
``` {r}
### set up vector of a values and initial param vals
a_vec <- seq(0, 1, .25)
t_vec <- 1:150
p_0   <- .15

params <- c(c_i = 0.1, e = 0.02, D = 0.5, a = 0)

### Run ODE on the whole shebang
p_list2 <- lapply(a_vec, function(a) {
  # a <- a_vec[2]
  params$a <- a
  p_vec <- ode(y = p_0, times = t_vec, func = lande_a_cco, parms = params) %>%
    as.data.frame() %>%
    setNames(c('t', 'p')) %>%
    mutate(a = a)
})

p_df2 <- bind_rows(p_list2)

ggplot(p_df2, aes(x = t, y = p, color = a)) +
  geom_line(aes(group = a)) +
  theme_minimal() +
  labs(x = 'Time',
       y = 'Proportion of populated sites')

```

At $a = 0$, rescue effect is non-existent, so should look like standard Lande with $D = 0.5$ (from part a) when all parameters are the same, and higher when $a > 0$.  Rerun with $c_i = 0.1, e = 0.01$ as in the first sim.

``` {r}
### set up vector of a values and initial param vals
a_vec <- seq(0, 1, .25)
t_vec <- 1:150
p_0   <- .15

params <- c(c_i = 0.1, e = 0.01, D = 0.5, a = 0)

### Run ODE on the whole shebang
p_list3 <- lapply(a_vec, function(a) {
  # a <- a_vec[2]
  params$a <- a
  p_vec <- ode(y = p_0, times = t_vec, func = lande_a_cco, parms = params) %>%
    as.data.frame() %>%
    setNames(c('t', 'p')) %>%
    mutate(a = a)
})

p_df3 <- bind_rows(p_list3)

ggplot(p_df3, aes(x = t, y = p, color = a)) +
  geom_line(data = p_df %>% filter(D == .5), 
            color = 'red4', alpha = .5, size = 3) +
  geom_line(aes(group = a)) +
  theme_minimal() +
  labs(x = 'Time',
       y = 'Proportion of populated sites')

```
