---
title: 'Primer of Ecology with $\textsf{R}$: chapter 4 Problems (Lumpies)'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    toc: false
    code_folding: hide
    number_sections: false
    theme: cerulean
    highlight: haddock
  pdf_document:
    toc: false
---

``` {r setup, echo = TRUE, message = FALSE, warning = FALSE}

knitr::opts_chunk$set(fig.width = 6, fig.height = 4, # fig.path = 'Figs/',
                      echo = TRUE, message = FALSE, warning = FALSE)

library(tidyverse)
library(RColorBrewer)

### The book has an R package! 
library(primer) # install.packages('primer')

```

# Chapter 4: Populations in Space{.tabset}

## 4.1. Equilibria

Derive expressions and calculate equilibria for the following metapopulation models, with $c_i = 0.05$, $e = 0.01$. Show your work — start with the differential equations, set to zero, and solve $p^{∗}$; then substitute in values for $c_i, e$.

(a) Levins model.
(b) Propagule rain model (`gotelli`).
(c) Propagule rain model that also includes both external and internal propagule production and dispersal.
(d) Hanski model.
(e) Lande (habitat destruction) model (with `D = 0.1`).

### Molly's attempts:
*currently on paper... will see if I have time to practice my LaTeX :)*

## 4.2. Habitat destruction

Compare different levels of habitat destruction.

(a) Use the habitat destruction model (`lande`) to compare 9 levels of destruction (`ds <- seq(0,.8, by=.1)`), using $c_i = 0.1, e = 0.01$. Plot of graph of the dynamics through time, and calculate the equilibria directly.
(b) Write an ODE function for a habitat destruction model with rescue effect. Let the “rescue” have an additional parameter, $a$, such that extinction rate is $ep(1 − ap)$.
(c) Let $D = 0.5, c_i = 0.1, e = 0.02$, and vary $a$ over five levels (including $a = 0, 1$) to investigate the effects of “relative rescue effect” on the equilibria and dynamics of a metapopulation.

### Molly's attempts:
(a) Use the Lande model to compare 9 levels of destruction (`ds <- seq(0,.8, by=.1)`), using $c_i = 0.1, e = 0.01$. Graph the dynamics through time.
```{r}
### copied pretty much exactly from the text...
library(deSolve)
prmsD <- c(ci = 0.1, e = 0.01, D = 0)
Ds <- seq(0,.8, by=.1)
Initial.p <- 0.01
t <- 1:200

lande <- function(t, y, parms) {
  p <- y[1]
  with(as.list(parms), {
    dp <- ci * p * (1 - D - p) - e * p
    return(list(dp))
  })
}

ps <- sapply(Ds, function(d) {
  prmsD["D"] <- d
  ode(y = Initial.p, times = t, func = lande, parms = prmsD)[, 2]
})

matplot(t, ps, type = "l", ylab = "p", xlab = "time")
text(c(200, 200, 200), ps[200, ], paste("D = ", Ds, sep = ""),
     adj = c(1, 0))
```

(b) Write an ODE function for a habitat destruction model with rescue effect. Let the “rescue” have an additional parameter, $a$, such that extinction rate is $ep(1 − ap)$.
```{r}
lande_rescue <- function(t, y, parms) {
  p <- y[1]
  with(as.list(parms), {
    dp <- ci * p * (1 - D - p) - e * p * (1 - a * p)
    return(list(dp))
  })
}
```

(c) Let $D = 0.5, c_i = 0.1, e = 0.02$, and vary $a$ over five levels (including $a = 0, 1$) to investigate the effects of “relative rescue effect” on the equilibria and dynamics of a metapopulation.
```{r}
prmsA <- c(ci = 0.1, e = 0.02, D = 0.5, a = 0)
as <- seq(0, 1, by = .25)
Initial.p <- 0.01
t <- 1:200

ps2 <- sapply(As, function(a) {
  prmsA["a"] <- a
  ode(y = Initial.p, times = t, func = lande_rescue, parms = prmsA)[, 2]
})

matplot(t, ps2, type = "l", ylab = "p", xlab = "time")
text(c(200, 200, 200), ps[200, ], paste("D = ", Ds, sep = ""),
     adj = c(1, 0))
```

