---
title: 'Ecological Dynamics ch 7'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    toc: false
    code_folding: hide
    number_sections: false
    theme: cerulean
    highlight: haddock
  pdf_document:
    toc: false
---

```{r setup, echo = FALSE, message = FALSE, warning = FALSE}

knitr::opts_chunk$set(fig.width = 6, fig.height = 4, # fig.path = 'Figs/',
                      echo = TRUE, message = FALSE, warning = FALSE)

source('https://raw.githubusercontent.com/oharac/src/master/R/common.R')
library(bbmle)
library(deSolve)

```

# Chapter 7: Ecosystems  {.tabset}

## 7.2.1 Constant production: linear functional response {.tabset}

### One-level system

This system w single functional group, "plants."  Carbon biomass as currency.  Carbon biomass density as $P(t)$ gC/m^2^.  Photosynthesis produces new biomass at $\phi$ gC/m^2^/day, and plant of mass $w$ loses mass (mortality and respiration) at $\delta_p w$ gC/day.

$$\frac{dP}{dt} = \phi - \delta_p P$$
So, steady state stock of plants at 
$$0 = \phi - \delta_p P \Longrightarrow P^* = \phi / \delta_p$$

### Two-level system

Add a second group, "herbivores," to the ecosystem.  Carbon biomass density of herbivores is $H(t)$ gC/m^2^; respiration & mortality loss at $\delta_h$ day^-1^, and linear functional response attack rate $\alpha_h$ m^2/day/gC.

$$\frac{dP}{dt} = \phi - \delta_p P - \alpha_h PH$$ 
and
$$\frac{dH}{dt} = \alpha_h PH - \delta_h H$$
Can find two steady states, one with $H=0$ and $P$ given by one-level steady state above.  The other is:
$$P^* = \delta_h / \alpha_h, \quad H^* = \frac{1}{\delta_h}(\phi - \delta_p P^*)$$

#### Function and params for two-level system 

This will calculate $\frac{dH}{dt}$ and $\frac{dP}{dt}$.

``` {r}
foodchain_2level <- function(t, y, params) {
  P <- y[1]
  H <- y[2]
  with(as.list(params), {
    dP_dt <- phi - d_p * P - a_h * P * H
    dH_dt <- a_h * P * H - d_h * H
    return(list(c(dP_dt, dH_dt)))
  })
}

```

#### Two-level dynamics

```{r}

### params for simulations

a_h <- 0.01 ### herbivore attack rate
phi <- 10 ### photosynthesis rate gC/m^2/day
d_h <- 0.5 ### carbon mass losses for herbivores
d_p <- 0.1 ### carbon mass losses for plants

pops_0 <- c(P = 25, H = 5) ### initial carbon densities of herbivore and plant
prms <- c(a_h = a_h, phi = phi, d_h = d_h, d_p = d_p)
t <- seq(0, 100, by = 1)
foodchain2_out <- ode(pops_0, t, foodchain_2level, prms)

foodchain2_df <- foodchain2_out %>%
  as.data.frame() %>%
  setNames(c('t', 'P', 'H'))
  
```

``` {r plot dynamics}
### Equilibrium
P_star2 <- d_h / a_h
H_star2 <- 1/d_h * (phi - d_p * P_star2)

### Isoclines
P_iso2 <- function(p) {
  h = (phi - d_p * p) / (a_h * p)
}
H_iso2 <- P_star2


### plot
ggplot(data = foodchain2_df, aes(x = P, y = H)) +
  ggtheme_plot() +
  stat_function(aes(x = P), fun = P_iso2, 
                linetype = 'dashed', color = 'darkred') +
  geom_vline(xintercept = H_iso2, 
             linetype = 'dashed', color = 'darkblue') +
  geom_segment(aes(xend = lead(P), yend = lead(H)),
               arrow = arrow(length = unit(0.07, 'inches')), 
               size = .25) +
  geom_point(x = P_star2, y = H_star2, size = 3) +
  geom_hline(yintercept = 0) + geom_vline(xintercept = 0) +
  ylim(c(0, max(foodchain2_df$H)))
```

#### Stability analysis using Jacobian (from Primer)

This is not in Ecological Dynamics, but based on Primer.

From the equil formulas above, $P^* = \frac{\delta_h}{\alpha_h}$ = `r P_star2`, $H^* = \frac{1}{\delta_h}(\phi - \delta_p P^*)$ = `r H_star2`

##### Creating, solving and using the Jacobian matrix

To know how each population growth rate changes in response to changes in the abundance each of the other population, we take the partial derivatives. 

$$\mathbf J = \begin{pmatrix}
    \frac{\partial \dot P}{\partial P} &
    \frac{\partial \dot P}{\partial H} \\
    \frac{\partial \dot H}{\partial P} &
    \frac{\partial \dot H}{\partial H}
  \end{pmatrix} =
  \begin{pmatrix}
    -\delta_p - \alpha_h H   & -\alpha_h P    \\
    \alpha_h H               & \alpha_h P - \delta_h
  \end{pmatrix}$$

We can replace the $P$ and $H$ in the Jacobian with the equilibria found above $P^*, H^*$.

$$\begin{pmatrix}
    -\delta_p - \alpha_h \frac{1}{\delta_h}(\phi - \delta_p \frac{\delta_h}{\alpha_h})
      & -\alpha_h \frac{\delta_h}{\alpha_h}    \\
    \alpha_h \frac{1}{\delta_h}(\phi - \delta_p \frac{\delta_h}{\alpha_h})
      & \alpha_h \frac{\delta_h}{\alpha_h} - \delta_h
  \end{pmatrix} = 
  \begin{pmatrix}
    \frac{-\alpha_h \phi}{\delta_h}        
      & -\delta_h\\
    \frac{\alpha_h \phi}{\delta_h} - \delta_p 
      & 0
  \end{pmatrix}$$

Recall the Routh-Hurwitz criterion:

$$ J_{11} + J_{22} < 0$$
$J_{11}$ is negative, so the plant population will self regulate, while $J_{22}$ is zero, so the herbivore population does not have a density-dependence effect on itself. ???

The other part of the Routh-Hurwitz criterion is the condition:

$$ J_{11}J_{22} - J_{12}J_{21} > 0$$
In the plant-herbivore context, this suggests that the plant declines due to the herbivore ($J_{12} < 0$) and the herbivore increases due to the plant ($J_{21} > 0$ as long as $\phi > \delta_p \frac{\delta_h}{\alpha_h}$). The signs of these elements make their product negative, and help make the above condition true.

Note: the $\phi  > \frac{\delta_p \delta_h}{\alpha_h}$ term indicates: primary production $\phi$ is high enough to support respiration/mortality losses from the plant population at the density needed to bring the herbivore production and loss into balance, or alternately, there's only a finite herbivore standing stock if the plant standing stock in the absennce of herbivory is larger than that required to sustain steady-state herbivore standing stock.

```{r}
# We can perform eigenanalysis given the parameters above.
Jac <- matrix(c(-a_h * phi / d_h, -d_h, a_h * phi / d_h - d_p,  0), 
              byrow = TRUE, nr = 2)
eig <- round(eigen(Jac)[["values"]],3)
```

If we performed eigenanalysis on the above Jacobian matrix, we would find that the eigenvalues are complex (`r eig`), meaning that the populations will oscillate or cycle, with period $\frac{2\pi}{\omega}$. 
* If the real parts are zero, the system will exhibit neutral stability
* If the real parts are negative, the system will converge toward the equilibrium.
* If the real parts are positive, the system will approach a stable oscillation away from the equilibrium.

### Three-level system

Add a third group, "consumers," which only eat herbivores, to the ecosystem.  Carbon biomass density of consumers is $C(t)$ gC/m^2^; respiration & mortality loss at $\delta_c$ day^-1^, and linear functional response attack rate $\alpha_c$ m^2/day/gC.  The system of equations:

$$\frac{dP}{dt} = \phi - \delta_p P - \alpha_h PH$$ 
$$\frac{dH}{dt} = \alpha_h PH - \delta_h H - \alpha_c H C$$
$$\frac{dC}{dt} = \alpha_c HC - \delta_c C$$
Can find three steady states:

* Plants only, with $C = H = 0$ and $P$ given by one-level steady state above.
* No consumers, with $C = 0$ and the system devolves to a two-level system
* Coexistence steady state, with
$$P^* = \frac{\phi}{\delta_p + \alpha_h H^*}, \quad H^* = \delta_c / \alpha_c, \quad C^* = \frac{1}{\delta_c}(\phi - \delta_p P^* - \delta_h H^*)$$

#### Function and params for three-level system 

This will calculate $\frac{dC}{dt}, \frac{dH}{dt}, \frac{dP}{dt}$.

``` {r} 
foodchain_3level <- function(t, y, params) {
  P <- y[1]
  H <- y[2]
  C <- y[3]
  with(as.list(params), {
    dP_dt <- phi - d_p * P - a_h * P * H
    dH_dt <- a_h * P * H - d_h * H - a_c * H * C
    dC_dt <- a_c * H * C - d_c * C
    return(list(c(dP_dt, dH_dt, dC_dt)))
  })
}

```

#### Three-level dynamics

To have sensible equilibria, the following needs to hold true:
$$\phi \geq \delta_p \left(\frac{\delta_h}{\alpha_h}\right) + \delta_h \left(\frac{\delta_c}{\alpha_c} \right)$$
```{r}

### params for simulations: what are some sensible numbers?
a_h <- 0.005 ### herbivore attack rate
a_c <- 0.015 ### consumer attack rate
phi <- 150   ### photosynthesis rate gC/m^2/day
d_h <- 0.90  ### carbon mass losses for herbivores
d_p <- 0.20  ### carbon mass losses for plants
d_c <- 0.93  ### carbon mass losses for consumers

phi_floor <- d_p * d_h / a_h + d_h * d_c / a_c
if(phi < phi_floor) {
  stop('phi value is too low')
}
  
### Equilibrium
H_star3 <- d_c / a_c
P_star3 <- phi / (d_p + a_h * H_star3)
C_star3 <- (phi - d_p * P_star3 - d_h * H_star3) / d_c
if(any(c(P_star3 < 0, H_star3 < 0, C_star3 < 0))) {
  stop('Param values lead to non-sensible equilibria')
}
```

``` {r run ode 3 level}
pops_0 <- c(P = 100, H = 100, C = 5) 
  ### initial carbon densities of herbivore, plant, and consumer
prms <- c(a_h = a_h, a_c = a_c, phi = phi, 
          d_h = d_h, d_p = d_p, d_c = d_c)
t <- seq(0, 100, by = 0.1)

foodchain3_out <- ode(pops_0, t, foodchain_3level, prms)

foodchain3_df <- foodchain3_out %>%
  as.data.frame() %>%
  setNames(c('t', 'P', 'H', 'C'))
  
```

``` {r plot dynamics 3 level pops over time}

equil_df <- data.frame(P = P_star3, H = H_star3, C = C_star3) %>%
  gather(level, pop) %>%
  mutate(level = fct_inorder(level))

pop_v_time_df <- foodchain3_df %>%
  gather(level, pop, -t) %>%
  mutate(level = fct_inorder(level))

ggplot(pop_v_time_df, aes(x = t, y = pop, color = level)) +
  ggtheme_plot() +
  geom_line() +
  geom_hline(data = equil_df, aes(yintercept = pop, color = level),
             linetype = 'dashed', size = .25) +
  geom_hline(yintercept = 0) +
  scale_color_manual(values = c('darkgreen', 'blue', 'darkred'))

```

``` {r plot dynamics 3 level phase plane}

library(plotly)

p <- plot_ly(data = foodchain3_df, 
             x = ~ P, y = ~ H, z = ~ C, 
             type = 'scatter3d', mode = 'lines',
             opacity = 1)

p

```

#### Stability analysis using Jacobian

From the equil formulas above:

* $P^* = \frac{\phi}{\delta_p + \alpha_h H^*}$ = `r P_star3`
* $H^* = \frac{\delta_c}{\alpha_c}$ = `r H_star3`
* $C^* = \frac{1}{\delta_c}(\phi - \delta_p P^* - \delta_h H^*)$ = `r C_star3`.

##### Creating, solving and using the Jacobian matrix

To know how each population growth rate changes in response to changes in the abundance each of the other population, we take the partial derivatives. 

$$\mathbf J = \begin{pmatrix}
    \frac{\partial \dot P}{\partial P} &
    \frac{\partial \dot P}{\partial H} &
    \frac{\partial \dot P}{\partial C} \\
    \frac{\partial \dot H}{\partial P} &
    \frac{\partial \dot H}{\partial H} &
    \frac{\partial \dot H}{\partial C} \\
    \frac{\partial \dot C}{\partial P} &
    \frac{\partial \dot C}{\partial H} &
    \frac{\partial \dot C}{\partial C}
  \end{pmatrix} =
  \begin{pmatrix}
    -\delta_p - \alpha_h H   
      & -\alpha_h P    
      & 0\\
    \alpha_h H               
      & \alpha_h P - \delta_h - \alpha_c C 
      & -\alpha_c H\\
    0
      & \alpha_c C
      & \alpha_c H - \delta_c
  \end{pmatrix}$$

We can replace the $P, H, C$ in the Jacobian with the equilibria found above $P^*, H^*, C^*$.

$$\mathbf J =   \begin{pmatrix}
    -\delta_p -  \frac{\alpha_h \delta_c}{\alpha_c}   
      & - \frac{\alpha_h \phi}{\delta_p + \frac{\alpha_h \delta_c}{\alpha_c}}    
      & 0\\
     \frac{\alpha_h \delta_c}{\alpha_c}               
      & \frac{\alpha_h \phi}{\delta_p + \alpha_h \frac{\delta_c}{\alpha_c}} 
          - \delta_h 
          - \frac{\alpha_c }{\delta_c}(\phi -  \frac{\delta_p \phi}{\delta_p + \frac{\alpha_h \delta_c}{\alpha_c}} 
          - \frac{\delta_h\delta_c}{\alpha_c}) 
      & -\frac{\alpha_c \delta_c}{\alpha_c}\\
    0
      & \frac{\alpha_c}{\delta_c}(\phi - \frac{\delta_p \phi}{\delta_p + \frac{\alpha_h \delta_c}{\alpha_c}} 
          - \frac{\delta_h \delta_c}{\alpha_c})
      & \frac{\alpha_c \delta_c}{\alpha_c} - \delta_c
  \end{pmatrix}$$
$$\Longrightarrow \mathbf J = \begin{pmatrix}
    -\delta_p -  \frac{\alpha_h\delta_c}{\alpha_c}   
      & - \frac{\alpha_h \phi}{\delta_p + \frac{\alpha_h \delta_c}{\alpha_c}}    
      & 0\\
    \frac{\alpha_h \delta_c}{\alpha_c}               
      & 0
      & -\delta_c\\
    0
      &   \frac{\alpha_h \phi}{\delta_p + \frac{\alpha_h \delta_c}{\alpha_c}} - \delta_h
      & 0
  \end{pmatrix}$$

Recall the Routh-Hurwitz criterion, assuming we can update simply for three species?:

$$ J_{11} + J_{22} + J_{33}< 0$$
$J_{11}$ is negative, so the plant population will self regulate, while $J_{22}$ and $J_{33}$ are zero, so the herbivore and consumer populations do not have a density-dependence effect on themselves. ???

The other part of the Routh-Hurwitz criterion is the condition (how to update for three species?):

$$ J_{11}J_{22} - J_{12}J_{21} > 0$$


```{r}
# We can perform eigenanalysis given the parameters above.
Jac3 <- matrix(
  c(-d_p - a_h*d_c / a_c,            - (a_h*phi) * (d_p + a_h*d_c / a_c),      0,
           a_h*d_c / a_c,                                              0,   -d_c,
                       0,    (a_h * phi) * (d_p + (a_h*d_c) / a_c) - d_h,     0), 
              byrow = TRUE, nr = 3)
eig3 <- round(eigen(Jac3)[["values"]],3)
```

When we performed eigenanalysis on the above Jacobian matrix, we find that the eigenvalues are complex (`r eig3`), meaning that the populations will oscillate or cycle.  What does it mean that one eigenvalue is not complex?
* If the real parts are zero, the system will exhibit neutral stability
* If the real parts are negative, the system will converge toward the equilibrium.
* If the real parts are positive, the system will approach a stable oscillation away from the equilibrium.

### Four-level system

Add a fourth group, "top predators," which only eat consumers, to the ecosystem.  Carbon biomass density of top predators is $T(t)$ gC/m^2^; respiration & mortality loss at $\delta_t$ day^-1^, and linear functional response attack rate $\alpha_t$ m^2/day/gC.  The system of equations:

$$\frac{dP}{dt} = \phi - \delta_p P - \alpha_h PH$$ 
$$\frac{dH}{dt} = \alpha_h PH - \delta_h H - \alpha_c H C$$
$$\frac{dC}{dt} = \alpha_c HC - \delta_c C - \alpha_c CT$$
$$\frac{dT}{dt} = \alpha_t CT - \delta_t T$$
Can find four steady states:

* Plants only, with $T = C = H = 0$ and $P$ given by one-level steady state above.
* No consumers, with $T = C = 0$ and the system devolves to a two-level system
* Steady state with no top predators, as in three-level system.
* Coexistence of all four levels - see Table 7.1.

#### Function and params for four-level system 

This will calculate $\frac{dC}{dt}, \frac{dH}{dt}, \frac{dP}{dt}$.

``` {r} 
foodchain_4level <- function(t, y, params) {
  P <- y[1]
  H <- y[2]
  C <- y[3]
  TT <- y[4]
  with(as.list(params), {
    dP_dt <- phi - d_p * P - a_h * P * H
    dH_dt <- a_h * P * H - d_h * H - a_c * H * C
    dC_dt <- a_c * H * C - d_c * C - a_t * C * TT 
    dT_dt <- a_t * C * TT - d_t * TT
    return(list(c(dP_dt, dH_dt, dC_dt, dT_dt)))
  })
}

```

#### Four-level dynamics

To have sensible equilibria, the following needs to hold true: ??? update
$$\phi \geq \delta_p \left(\frac{\delta_h}{\alpha_h}\right) + \delta_h \left(\frac{\delta_c}{\alpha_c} \right)$$
```{r}

### params for simulations: what are some sensible numbers?
a_h <- 0.015 ### herbivore attack rate
a_c <- 0.025 ### consumer attack rate
a_t <- 0.035 ### top predator attack rate
phi <- 150   ### photosynthesis rate gC/m^2/day
d_h <- 0.90  ### carbon mass losses for herbivores
d_p <- 0.25  ### carbon mass losses for plants
d_c <- 0.90  ### carbon mass losses for consumers
d_t <- 0.90  ### carbon mass losses for top predators

phi_floor <- d_p * d_h / a_h + d_h * d_c / a_c
if(phi < phi_floor) {
  stop('phi value is too low')
}
  
### Equilibrium
C_star4 <- d_t / a_t
P_star4 <- (d_h + a_c * C_star4) / a_h
H_star4 <- (phi - d_p * P_star4) / (a_h * P_star4)
T_star4 <- (a_c * H_star4 - d_c) / (a_t)
if(any(c(P_star4 < 0, H_star4 < 0, C_star4 < 0, T_star4 < 0))) {
  stop('Param values lead to non-sensible equilibria')
}
```

``` {r run ode 4 level}
pops_0 <- c(P = 100, H = 100, C = 20, TT = 10) 
  ### initial carbon densities of herbivore, plant, and consumer
prms4 <- c(a_h = a_h, a_c = a_c, a_t = a_t, phi = phi, 
          d_h = d_h, d_p = d_p, d_c = d_c, d_t = d_t)
t <- seq(0, 100, by = 0.1)

foodchain4_out <- ode(pops_0, t, foodchain_4level, prms4)

foodchain4_df <- foodchain4_out %>%
  as.data.frame() %>%
  setNames(c('t', 'P', 'H', 'C', 'T'))
  
```

``` {r plot dynamics 4 level pops over time}

equil_df <- data.frame(P = P_star4, H = H_star4, C = C_star4, T = T_star4) %>%
  gather(level, pop) %>%
  mutate(level = fct_inorder(level))

pop_v_time_df <- foodchain4_df %>%
  gather(level, pop, -t) %>%
  mutate(level = fct_inorder(level))

ggplot(pop_v_time_df, aes(x = t, y = pop, color = level)) +
  ggtheme_plot() +
  geom_line() +
  geom_hline(data = equil_df, aes(yintercept = pop, color = level),
             linetype = 'dashed', size = .25) +
  geom_hline(yintercept = 0) +
  scale_color_manual(values = c('darkgreen', 'blue', 'red', 'darkred'))

```

``` {r plot dynamics 4 level phase plane}

library(plotly)

p <- plot_ly(data = foodchain4_df %>%
               mutate(color = as.factor(T)), 
             x = ~ P, y = ~ H, z = ~ C,
             type = 'scatter3d', mode = 'lines',
             line = list(width = 3, color = ~ color, reverscale = FALSE),
             opacity = 1)

p

```


## 7.2.2 Logistic primary production {.tabset}

### One-level system

Modify plant carbon biomass to grow logistically to carrying capacity $K$:

$$\frac{dP}{dt} = rP \left( 1 - \frac{P}{K} \right)$$

### Two-level system

Add a second group, "herbivores," to the ecosystem, following assumptions in 7.2.1.

$$\frac{dP}{dt} = rP \left( 1 - \frac{P}{K} \right) - \alpha_h P H$$ 
and
$$\frac{dH}{dt} = \alpha_h PH - \delta_h H$$
Can find two steady states, one with $H=0$ and $P = K$.  The other is:
$$P^* = \delta_h / \alpha_h, \quad H^* = \frac{r}{\alpha_h} \left(1 - \frac{P^*}{K} \right)$$

#### Function and params for two-level system 

``` {r}
foodchain_2level <- function(t, y, params) {
  P <- y[1]
  H <- y[2]
  with(as.list(params), {
    dP_dt <- r_p * P * (1 - P / K_p) - a_h * H * P
    dH_dt <- a_h * P * H - d_h * H
    return(list(c(dP_dt, dH_dt)))
  })
}
```

#### Two-level dynamics

```{r}

### params for simulations
a_h <- 0.01 ### herbivore attack rate
r_p <- 0.5 ### plant intrinsic growth rate
d_h <- 0.5 ### carbon mass losses for herbivores
K_p <- 200 ### carrying capacity for plants

pops_0 <- c(P = 25, H = 5) ### initial carbon densities of herbivore and plant
prms <- c(a_h = a_h, r_p = r_p, d_h = d_h, K_p = K_p)
t <- seq(0, 100, by = 1)
foodchain2_out <- ode(pops_0, t, foodchain_2level, prms)

foodchain2_df <- foodchain2_out %>%
  as.data.frame() %>%
  setNames(c('t', 'P', 'H'))
  
```

``` {r plot dynamics 2}
### Equilibrium
P_star2 <- d_h / a_h
H_star2 <- r_p / a_h * (1 - P_star2 / K_p)

### Isoclines
P_iso2 <- function(p) {
  h <- r_p / a_h * (1 - p / K_p)
}
H_iso2 <- P_star2


### plot
ggplot(data = foodchain2_df, aes(x = P, y = H)) +
  ggtheme_plot() +
  stat_function(aes(x = P), fun = P_iso2, 
                linetype = 'dashed', color = 'darkred') +
  geom_vline(xintercept = H_iso2, 
             linetype = 'dashed', color = 'darkblue') +
  geom_segment(aes(xend = lead(P), yend = lead(H)),
               arrow = arrow(length = unit(0.07, 'inches')), 
               size = .25) +
  geom_point(x = P_star2, y = H_star2, size = 3) +
  geom_hline(yintercept = 0) + geom_vline(xintercept = 0) +
  ylim(c(0, max(foodchain2_df$H)))
```

#### Stability analysis using Jacobian (from Primer)

From the equil formulas above, $P^* = \frac{\delta_h}{\alpha_h}$ = `r P_star2`, $H^* = \frac{r}{\alpha_h} \left(1 - \frac{P^*}{K} \right$ = `r H_star2`

##### Creating, solving and using the Jacobian matrix

To know how each population growth rate changes in response to changes in the abundance each of the other population, we take the partial derivatives. 

$$\mathbf J = \begin{pmatrix}
    \frac{\partial \dot P}{\partial P} &
    \frac{\partial \dot P}{\partial H} \\
    \frac{\partial \dot H}{\partial P} &
    \frac{\partial \dot H}{\partial H}
  \end{pmatrix} =
  \begin{pmatrix}
    r \left(1 - \frac{2P}{K} \right) - \alpha_h H  
      & -\alpha_h P    \\
    \alpha_h H               
      & \alpha_h P - \delta_h
  \end{pmatrix}$$

We can replace the $P$ and $H$ in the Jacobian with the equilibria found above $P^*, H^*$.

$$\begin{pmatrix}
    r \left(1 - \frac{2\frac{\delta_h}{\alpha_h}}{K} \right) - \alpha_h \frac{r}{\alpha_h} \left(1 - \frac{\frac{\delta_h}{\alpha_h}}{K} \right)  & -\alpha_h \frac{\delta_h}{\alpha_h}    \\
    \alpha_h \frac{r}{\alpha_h} \left(1 - \frac{\frac{\delta_h}{\alpha_h}}{K} \right)               & \alpha_h \frac{\delta_h}{\alpha_h} - \delta_h
  \end{pmatrix} = 
  \begin{pmatrix}
    \frac{-r \delta_h}{\alpha_h K}        
      & -\delta_h\\
    r \left(1 - \frac{\delta_h}{\alpha_h K} \right)
      & 0
  \end{pmatrix}$$

Recall the Routh-Hurwitz criterion:

$$ J_{11} + J_{22} < 0$$
$J_{11}$ is negative, so the plant population will self regulate, while $J_{22}$ is zero, so the herbivore population does not have a direct density-dependence effect on itself.

$$ J_{11}J_{22} - J_{12}J_{21} > 0$$
$J_{11}J_{22} = 0$, so stability depends on $J_{12}J_{21} < 0$. The plant declines due to the herbivore ($J_{12} < 0$) and the herbivore increases due to the plant ($J_{21} > 0$ as long as $P^* < K$).

$$\begin{pmatrix}
    \frac{-r \delta_h}{\alpha_h K}        
      & -\delta_h\\
    r \left(1 - \frac{\delta_h}{\alpha_h K} \right)
      & 0
  \end{pmatrix}$$
```{r}
# We can perform eigenanalysis given the parameters above.
Jac <- matrix(c(-r_p * d_h / (a_h * K_p), -d_h, r_p * (1 - d_h / (a_h * K_p)), 0), 
              byrow = TRUE, nr = 2)
eig <- round(eigen(Jac)[["values"]],3)
```

If we performed eigenanalysis on the above Jacobian matrix, we would find that the eigenvalues are complex (`r eig`), meaning that the populations will oscillate or cycle, with period $\frac{2\pi}{\omega}$. 
* If the real parts are zero, the system will exhibit neutral stability
* If the real parts are negative, the system will converge toward the equilibrium.
* If the real parts are positive, the system will approach a stable oscillation away from the equilibrium.

### Three-level system

Add a third group, "consumer," which only eat herbivores, to the ecosystem.  Carbon biomass density of consumers is $C(t)$ gC/m^2^; respiration & mortality loss at $\delta_c$ day^-1^, and linear functional response attack rate $\alpha_c$ m^2/day/gC.  The system of equations:

$$\frac{dP}{dt} = rP \left( 1 - \frac{P}{K} \right) - \alpha_h PH$$ 
$$\frac{dH}{dt} = \alpha_h PH - \delta_h H - \alpha_c H C$$
$$\frac{dC}{dt} = \alpha_c HC - \delta_c C$$
Can find three steady states:

* Plants only, with $C = H = 0$ and $P = K$.
* No consumers, with $C = 0$ and the system devolves to a two-level system
* Coexistence steady state, with
    * $P^* = K \left(1 - \frac{\alpha_h H^*}{r} \right)$
    * $H^* = \delta_c / \alpha_c$
    * $C^* = \frac{1}{\delta_c} \left[ rP^* \left( 1 - \frac{P^*}{K} \right) - \delta_h H^* \right]$

#### Function and params for three-level system 

This will calculate $\frac{dC}{dt}, \frac{dH}{dt}, \frac{dP}{dt}$.

``` {r} 
foodchain_3level <- function(t, y, params) {
  P <- y[1]
  H <- y[2]
  C <- y[3]
  with(as.list(params), {
    dP_dt <- r_p * P * (1 - P / K_p) - a_h * H * P
    dH_dt <- a_h * P * H - d_h * H - a_c * H * C
    dC_dt <- a_c * H * C - d_c * C
    return(list(c(dP_dt, dH_dt, dC_dt)))
  })
}
```

#### Three-level dynamics

To have sensible equilibria, the following needs to hold true:

$$K \left(1 - \frac{\alpha_h \delta_c}{\alpha_c r} \right) \geq \frac{\delta_h}{\alpha_h}$$

```{r}

### params for simulations: what are some sensible numbers?
a_h <- 0.005 ### herbivore attack rate
a_c <- 0.015 ### consumer attack rate
r_p <- 0.5  ### plant intrinsic growth rate
K_p <- 500   ### plant carrying capacity
d_h <- 0.90  ### carbon mass losses for herbivores
d_c <- 0.90  ### carbon mass losses for consumers

K_test <- K_p * (1 - (a_h * d_c) / (a_c * r_p)) >= d_h / a_h
if(!K_test) {
  stop('parameters are problematic')
}

### Equilibrium
H_star3 <- d_c / a_c
P_star3 <- K_p * (1 - a_h * H_star3 / r_p)
C_star3 <- (r_p * P_star3 * (1 - P_star3 / K_p) - d_h * H_star3) / d_c
if(any(c(P_star3 < 0, H_star3 < 0, C_star3 < 0))) {
  stop('Param values lead to non-sensible equilibria')
}
```

``` {r run ode 3 level 2}
pops_0 <- c(P = 100, H = 100, C = 5) 
  ### initial carbon densities of herbivore, plant, and consumer
prms <- c(a_h = a_h, a_c = a_c, r_p = r_p, 
          d_h = d_h, K_p = K_p, d_c = d_c)
t <- seq(0, 100, by = 0.1)

foodchain3_out <- ode(pops_0, t, foodchain_3level, prms)

foodchain3_df <- foodchain3_out %>%
  as.data.frame() %>%
  setNames(c('t', 'P', 'H', 'C'))
  
```

``` {r plot dynamics 3 level pops over time 2}

equil_df <- data.frame(P = P_star3, H = H_star3, C = C_star3) %>%
  gather(level, pop) %>%
  mutate(level = fct_inorder(level))

pop_v_time_df <- foodchain3_df %>%
  gather(level, pop, -t) %>%
  mutate(level = fct_inorder(level))

ggplot(pop_v_time_df, aes(x = t, y = pop, color = level)) +
  ggtheme_plot() +
  geom_line() +
  geom_hline(data = equil_df, aes(yintercept = pop, color = level),
             linetype = 'dashed', size = .25) +
  geom_hline(yintercept = 0) +
  scale_color_manual(values = c('darkgreen', 'blue', 'darkred'))

```

``` {r plot dynamics 3 level phase plane 2}

library(plotly)

p <- plot_ly(data = foodchain3_df, 
             x = ~ P, y = ~ H, z = ~ C, 
             type = 'scatter3d', mode = 'lines',
             opacity = 1)

p

```

### Four-level system (not correct?)

Add a fourth group, "top predators," which only eat consumers, to the ecosystem.  Carbon biomass density of top predators is $T(t)$ gC/m^2^; respiration & mortality loss at $\delta_t$ day^-1^, and linear functional response attack rate $\alpha_t$ m^2/day/gC.  The system of equations:

$$\frac{dP}{dt} = rP \left( 1 - \frac{P}{K} \right) - \alpha_h PH$$ 
$$\frac{dH}{dt} = \alpha_h PH - \delta_h H - \alpha_c H C$$
$$\frac{dC}{dt} = \alpha_c HC - \delta_c C - \alpha_c CT$$
$$\frac{dT}{dt} = \alpha_t CT - \delta_t T$$
Can find four steady states:

* Plants only, with $T = C = H = 0$ and $P$ given by one-level steady state above.
* No consumers, with $T = C = 0$ and the system devolves to a two-level system
* Steady state with no top predators, as in three-level system.
* Coexistence of all four levels:
    * $P^* = \frac{1}{\alpha_h}(\delta_h + \alpha_c C^*)$
    * $H^* = \frac{r}{\alpha_h}\left(1 - \frac{P}{K} \right)$
    * $C^* = \frac{\delta_t}{\alpha_t}$
    * $T^* = \frac{1}{\alpha_t}(\alpha_c H^* - \delta_c)$

#### Function and params for four-level system 

This will calculate $\frac{dC}{dt}, \frac{dH}{dt}, \frac{dP}{dt}$.

``` {r} 
foodchain_4level <- function(t, y, params) {
  P <- y[1]
  H <- y[2]
  C <- y[3]
  TT <- y[4]
  with(as.list(params), {
    dP_dt <- r_p * P * (1 - P / K_p) - a_h * H * P
    dH_dt <- a_h * P * H - d_h * H - a_c * H * C
    dC_dt <- a_c * H * C - d_c * C - a_t * C * TT 
    dT_dt <- a_t * C * TT - d_t * TT
    return(list(c(dP_dt, dH_dt, dC_dt, dT_dt)))
  })
}

```

#### Four-level dynamics??? not correct

To have sensible equilibria, the following needs to hold true: ??? update
$$\phi \geq \delta_p \left(\frac{\delta_h}{\alpha_h}\right) + \delta_h \left(\frac{\delta_c}{\alpha_c} \right)$$
```{r}

### params for simulations: what are some sensible numbers?
a_h <- 0.015 ### herbivore attack rate
a_c <- 0.025 ### consumer attack rate
a_t <- 0.035 ### top predator attack rate
r_p <- 0.25  ### intrinsic growth rate for plants
d_h <- 0.90  ### carbon mass losses for herbivores
K_p <- 500   ### carrying capacity for plants
d_c <- 0.90  ### carbon mass losses for consumers
d_t <- 0.90  ### carbon mass losses for top predators

  
### Equilibrium
C_star4 <- d_t / a_t
P_star4 <- (d_h + a_c * C_star4) / a_h
H_star4 <- r_p / a_h * (1 - P_star4 / K_p)
T_star4 <- (a_c * H_star4 - d_c) / (a_t)
# if(any(c(P_star4 < 0, H_star4 < 0, C_star4 < 0, T_star4 < 0))) {
#   stop('Param values lead to non-sensible equilibria')
# }
```

``` {r run ode 4 level 2}
pops_0 <- c(P = 100, H = 100, C = 20, TT = 10) 
  ### initial carbon densities of herbivore, plant, and consumer
prms4 <- c(a_h = a_h, a_c = a_c, a_t = a_t, r_p = r_p, 
           d_h = d_h, K_p = K_p, d_c = d_c, d_t = d_t)
t <- seq(0, 100, by = 0.1)

foodchain4_out <- ode(pops_0, t, foodchain_4level, prms4)

foodchain4_df <- foodchain4_out %>%
  as.data.frame() %>%
  setNames(c('t', 'P', 'H', 'C', 'T'))
  
```

``` {r plot dynamics 4 level pops over time 2}

equil_df <- data.frame(P = P_star4, H = H_star4, C = C_star4, T = T_star4) %>%
  gather(level, pop) %>%
  mutate(level = fct_inorder(level))

pop_v_time_df <- foodchain4_df %>%
  gather(level, pop, -t) %>%
  mutate(level = fct_inorder(level))

ggplot(pop_v_time_df, aes(x = t, y = pop, color = level)) +
  ggtheme_plot() +
  geom_line() +
  geom_hline(data = equil_df, aes(yintercept = pop, color = level),
             linetype = 'dashed', size = .25) +
  geom_hline(yintercept = 0) +
  scale_color_manual(values = c('darkgreen', 'blue', 'red', 'darkred'))

```


## 7.2.3 Type II functional response {.tabset}

### One-level system

Plant carbon biomass as constant primary production (same as 7.2.1)

$$\frac{dP}{dt} = \phi - \delta_p P$$

### Two-level system

Add a second group, "herbivores," to the ecosystem, but now the herbivores have a type II functional response.

$$\frac{dP}{dt} = \phi - \delta_p P - \frac{\alpha_h H P}{P + P_0}$$
and
$$\frac{dH}{dt} = \frac{\alpha_h PH}{P + P_0} - \delta_h H$$
Can find two steady states, one with $H=0$ and $P = \phi/\delta_p$.  The other is:

$$P^* = \frac{\delta_h P_0}{\alpha_h - \delta_h}, \quad H^* = \frac{\phi - \delta_p P^*}{\delta_h}$$

These should be positive, as long as:

$$\frac{\phi}{\delta_p} \geq \frac{\delta_h P_0}{\alpha_h - \delta_h}$$

#### Function and params for two-level system 

``` {r}
foodchain_2level <- function(t, y, params) {
  P <- y[1]
  H <- y[2]
  with(as.list(params), {
    dP_dt <- phi - d_p * P - (a_h * H * P) / (P + P_0)
    dH_dt <- (a_h * H * P) / (P + P_0) - d_h * H
    return(list(c(dP_dt, dH_dt)))
  })
}
```

#### Two-level dynamics

```{r}

### params for simulations
a_h <- 0.75 ### herbivore attack rate
phi <- 150   ### plant constant primary production
d_h <- 0.32  ### carbon mass losses for herbivores
d_p <- 0.25  ### carrying capacity for plants
P_0 <- 200

pops_0 <- c(P = 25, H = 5) ### initial carbon densities of herbivore and plant
prms <- c(a_h = a_h, phi = phi, d_h = d_h, d_p = d_p, P_0 = P_0)
t <- seq(0, 100, by = 1)
foodchain2_out <- ode(pops_0, t, foodchain_2level, prms)

foodchain2_df <- foodchain2_out %>%
  as.data.frame() %>%
  setNames(c('t', 'P', 'H'))
  
```

``` {r plot dynamics 3}
### Equilibrium
P_star2 <- d_h * P_0 / (a_h - d_h)
H_star2 <- (phi - d_p * P_star2) / d_h

### Isoclines
P_iso2 <- function(p) {
  h <- (phi - d_p * p) / (a_h * p) * (p + P_0)
}
H_iso2 <- d_h / a_h * P_0 / (1 - d_h / a_h)


### plot
ggplot(data = foodchain2_df, aes(x = P, y = H)) +
  ggtheme_plot() +
  stat_function(aes(x = P), fun = P_iso2, 
                linetype = 'dashed', color = 'darkred') +
  geom_vline(xintercept = H_iso2, 
             linetype = 'dashed', color = 'darkblue') +
  geom_segment(aes(xend = lead(P), yend = lead(H)),
               arrow = arrow(length = unit(0.07, 'inches')), 
               size = .25) +
  geom_point(x = P_star2, y = H_star2, size = 3) +
  geom_hline(yintercept = 0) + geom_vline(xintercept = 0) +
  ylim(c(0, max(foodchain2_df$H)))
```

### Three-level system

The "consumer" functional group now has a type II functional response as well.

* $\frac{dP}{dt} = \phi - \delta_p P - \frac{\alpha_h H P}{P + P_0}$
* $\frac{dH}{dt} = \frac{\alpha_h PH}{P + P_0} - \delta_h H - \frac{\alpha_c C H}{H + H_0}$
* $\frac{dC}{dt} = \frac{\alpha_c C H}{H + H_0} - \delta_c C$

As usual, can find three steady states; coexistence steady state, with

* $P^* = K \left(1 - \frac{\alpha_h H^*}{r} \right)$
* $H^* = \frac{\delta_c H_0}{\alpha_c - \delta_c}$
* $C^* = \frac{1}{\delta_c}(\phi - \delta_p P^* - \delta_h H^*)$

#### Function and params for three-level system 

This will calculate $\frac{dC}{dt}, \frac{dH}{dt}, \frac{dP}{dt}$.

``` {r} 
foodchain_3level <- function(t, y, params) {
  P <- y[1]
  H <- y[2]
  C <- y[3]
  with(as.list(params), {
    dP_dt <- phi - d_p * P - (a_h * H * P) / (P + P_0)
    dH_dt <- (a_h * H * P) / (P + P_0) - d_h * H - (a_c * C * H) / (H + H_0)
    dC_dt <- (a_c * C * H) / (H + H_0) - d_c * C
    return(list(c(dP_dt, dH_dt, dC_dt)))
  })
}
```

#### Three-level dynamics

To have sensible equilibria, the following needs to hold true:

$$K \left(1 - \frac{\alpha_h \delta_c}{\alpha_c r} \right) \geq \frac{\delta_h}{\alpha_h}$$

```{r}

### params for simulations: what are some sensible numbers?
a_h <- 0.75  ### herbivore attack rate
a_c <- 0.4  ### consumer attack rate
phi <- 150   ### plant constant primary production
d_h <- 0.32  ### carbon mass losses for herbivores
d_p <- 0.25  ### carrying capacity for plants
P_0 <- 200
H_0 <- 150
d_c <- 0.20  ### carbon mass losses for consumers

### Equilibrium
H_star3 <- (d_c * H_0) / (a_c - d_c)
### params for quadratic formula:
    a <- d_p
    b <- - (phi - d_p * P_0 - a_h * H_star3)
    c <- -phi * P_0
P_star3 <- (-b + sqrt(b^2 - 4 * a * c)) / (2 * a)
C_star3 <- (phi - d_p * P_star3 - d_h * H_star3) / d_c

if(any(c(P_star3 < 0, H_star3 < 0, C_star3 < 0))) {
  stop('Param values lead to non-sensible equilibria')
}
```

``` {r run ode 3 level 3}
pops_0 <- c(P = 100, H = 100, C = 5) 
  ### initial carbon densities of herbivore, plant, and consumer
prms <- c(a_h = a_h, a_c = a_c, phi = phi, 
          d_h = d_h, d_p = d_p, d_c = d_c, P_0 = P_0, H_0 = H_0)
t <- seq(0, 250, by = 0.1)

foodchain3_out <- ode(pops_0, t, foodchain_3level, prms)

foodchain3_df <- foodchain3_out %>%
  as.data.frame() %>%
  setNames(c('t', 'P', 'H', 'C'))
  
```

``` {r plot dynamics 3 level pops over time 3}

equil_df <- data.frame(P = P_star3, H = H_star3, C = C_star3) %>%
  gather(level, pop) %>%
  mutate(level = fct_inorder(level))

pop_v_time_df <- foodchain3_df %>%
  gather(level, pop, -t) %>%
  mutate(level = fct_inorder(level))

ggplot(pop_v_time_df, aes(x = t, y = pop, color = level)) +
  ggtheme_plot() +
  geom_line() +
  geom_hline(data = equil_df, aes(yintercept = pop, color = level),
             linetype = 'dashed', size = .25) +
  geom_hline(yintercept = 0) +
  scale_color_manual(values = c('darkgreen', 'blue', 'darkred'))

```

``` {r plot dynamics 3 level phase plane 3}

library(plotly)

p <- plot_ly(data = foodchain3_df, 
             x = ~ P, y = ~ H, z = ~ C, 
             type = 'scatter3d', mode = 'lines',
             opacity = 1)

p

```

