---
title: "R Practice"
author: "Sean Denny"
date: "7/8/2019"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Questions:

1. If I open RStudio, will all of the Github files already be updated/pulled? Or do I need to pull again?

2. Can I only pull if someone has changed something? Or should I be able to pull regardless?

3. What is the white M in the blue box next to our file name for teamNudi?

4. Confused about the notation for sapply (chunk 10).

5. What is RColorBrewer for?

6. For graphing a matrix,how do I change the scale on the right so it's not continuous but just reflects the three N0 values?

# Header
## Smaller
### This is as small as it gets, I think

I can now just type normally and it's the regular font size, but it's not bold. It's equivalent to size ###. If I wanted to bold it though, I could just **do this**. Or, I could italicize by doing *this*. If I wanted to make a list...

1. I
2. Could
3. Do
4. This

To add a new R chunk, use command+option+i, or just go to insert and insert an R script.

```{r}
mean <- 8
```


To reference something in the text in Rmarkdown use backticks and "r {function name}". Example, the mean is **`r {mean}`**. You can also use backticks just to write a function in Rmarkdown text without it applying the function. Example `\pagebreak`.

```{r Load tidyverse and RColorBrewer, message = FALSE}
# start by loading packages

library(tidyverse)
library(RColorBrewer)
```

```{r Load book package, message = FALSE, warning = FALSE}
#Use warning = FALSE in R chunk header to prevent Rmarkdown including warning messages in HTML or PDF outputs.

#This book has it's own R package

# install.packages('primer')
library(primer)
```

Example of using case_when to create a new column (with new values) based on criteria:

nc_new <- nc_births %>%
  mutate(
    sdp = case_when(
      smoke == 0 ~ "nonsmoker",
      smoke == 1 ~ "smoker"
    )
    
Using filter to select values that are everything but NA, or any specified value (the second option):

filter(!is.na(sdp)) *OR*
filter(sdp != "NA") 

**Finally**, if you wanted to do a pagebreak you can use this:

\pagebreak

# Chapter 1 {.tabset}

(By using {.tabset} in Rmarkdown, you can create tabs in the HTML format for the following headers, I believe.)

##1.3 Exploring Population Growth

Simple graphing of population size (Fig 1.3):

```{r}
# Original version

N <- c(1, 3, 9, 27, 81)
year <- 2001:2005
plot (year, N)
```

```{r}
# Sean's version

exp_growth <- data.frame(N = c(1, 3, 9, 27, 81), year = 2001:2005)

exp_growth_plot <- ggplot(exp_growth, aes(x= year, y = N)) +
   geom_point(color = 'red', size = 2) +
  theme_bw() +
  labs(x = 'Year', y = 'Population size / N')
exp_growth_plot
```

#### Vectorized math

Here we divide each element of one vector (the second through fifth element of $N$) by each element of another vector (the first through fourth elements of $N$).

```{r}
rates = N[2:5/1:4]
rates

rates = N[2:5]/N[1:4]
rates
# [1] 3 3 3 3

# Same rate of change for each population
```

### 1.3.1 Projecting population size

Here we calculate population sizes for 10 time points beyond the initial. First we assign values for $N_0$, $\lambda$, and time.

```{r}

N0 <- 1
lambda <- 2
time <- 0:10
```

Next we calculate $N_t$ directly using our general formula. 

``` {r}

Nt <- N0*lambda^time
Nt

# [1] 1 2 4 8 16 32 64 128 256 5121024

```

### 1.3.2 Effects of Initial Population Size

We first set up several different initial values, provide a fixed $\lambda$, and set times from zero to 4.

```{r}
N0 <- c(10, 20, 30)
lambda <- 2
time <- 0:4
```

We calculate population sizes at once using `sapply` to _apply_ a function (`n*lambda^time`) to each element of the first argument (each element of `N0`).

```{r}
Nt.s <- sapply(N0, function(n) n*lambda^time)
Nt.s

#     [,1] [,2] [,3]
# [1,]   10   20   30
# [2,]   20   40   60
# [3,]   40   80  120
# [4,]   80  160  240
# [5,]  160  320  480
```

The result is a matrix, and we see $N_0$ in the first row, and each population is in its own column. Note that population 2 is always twice as big as population 1.

#### Graphing a Matrix (Figs. 1.3a, 1.3b)

We can use matplot to plot a matrix vs. a single vector on the X-axis. By default it labels the points according to the number of the column

```{r}
matplot(time, Nt.s, pch = 1:3)
```

We can also plot it with a log scale on the y-axis. 
``` {r}
matplot(time, Nt.s, log = "y", pch = 1:3)
```

```{r}
#Sean's

# Use crossing() to make a data frame for all combinations of N0 and time
# Use mutate() to create a new column)

exp_pop_df <- crossing(time, N0) %>% 
  mutate(Nt = N0*lambda^time)

exp_pop_plot <- ggplot(exp_pop_df, aes(x = time, y = Nt, color = N0)) +
  geom_point() +
  labs(x = 'Year', y = 'Population size (Nt)') +
  geom_line(aes(group = N0)) +
  theme_bw()
exp_pop_plot

# Could use 'alpha =' in aes brackets to change transparent (from 0 to 1, I'm guessing.)

#Also, #scale_color_manual to change colors manually. How do I do this?
```

### 1.3.3 Effects of different per capita growth rates

Here we demonstrate the effects on growth of $\lambda > 1$ and $\lambda < 1$. We set $N_0 = 100$, and time, and then pick three different $\lambda$.
```{r}
N0 <- 100
time <- 0:3
lambdas <- c(0.5, 1, 1.5)
```

We use `sapply` again to apply the geometric growth function to each $\lambda$. This time, `x` stands for each $\lambda$, which our function then uses to calculate population size. We then plot it, and add a reference line and a little text.

```{r}
N.all <- sapply(lambdas, function(x) N0 * x^time)
matplot(time, N.all, xlab = "Years", ylab = "N", pch = 1:3)
abline(h = N0, lty = 3)
text(0.5, 250, expression(lambda > 1), cex = 1.2)
text(0.5, 20, expression(lambda < 1), cex = 1.2)
```

The reference line is a ***h***orizontal line with the ***l***ine ***ty***pe dotted. Our text simply indicates the regions of positive and negative growth.

### 1.3.4 Average growth rate

#### Comparing arithmetic and geometric averages (Fig. 1.5)

First we select the number of observed $R$ $(t = 5)$; this will require that we use six years of Song Sparrow data.

```{r}
t <- 5
data(sparrows)
SS6 <- sparrows[1:(t + 1), ]
```


Next we calculate $\lambda$ for each generation, from $t$ to $t + 1$, and calculate the arithmetic and geometric means.

```{r}
SSgr <- SS6$Count[2:(t + 1)]/SS6$Count[1:t]
lam.A <- sum(SSgr)/t
lam.G <- prod(SSgr)^(1/t)
```

Now we can plot the data, and the projections based on the two averages (Fig. 1.5).

```{r}
N0 <- SS6$Count[1]
plot(0:t, SS6$Count, ylab = "Projected Population Size")
lines(0:t, N0 * lam.A^(0:t), lty = 2)
lines(0:t, N0 * lam.G^(0:t), lty = 1)
legend(0, 70, 
       c("Arithmetic Ave.", "Geometric Ave."),
       title = "Projections Based On:", 
       lty = 2:1, bty = "n", xjust = 0)
```

## 1.4 Continuous Exponential Growth

### 1.4.1 Motivating continuous exponential growth

#### Numerical approximation of _e_

Here we use brute force to try to get an approximate solution to eq. 1.9. We’ll let n be the number of divisions within one year. This implies that the finite rate of increase during each of these fractional time steps is $r_d/n$. Let the $\lambda = 2$ and therefore $r_d = 1$. Note that because $N_0 = 1$, we could ignore it, but let’s keep it in for completeness.

```{r}
n <- 0:100
N0 <- 1
rd <- 1
```

Next, we calculate $\left( 1 + \frac{r_d}{n} \right)^n$ for ever larger values of $n$.
```{r}
N1 <- N0 * (1 + rd/n)^n
```

Last, we plot the ratio and add some fancy math text to the plot (see `?plotmath` for details on mathematical typesetting in R).

```{r}
plot(n, N1/N0, type = "l")
text(50, 2, "For n = 100,")
text(50, 1.6, 
     bquote((1+frac("r"["d"],"n"))^"n" == .(round(N1[101]/N0, + 3))))
```

Fig. 1.6: The limit to subdividing reproduction into smaller steps. We can compare this numerical approximation to the true value, $e^1 = 2.718$.

#### Projecting a continuous population

We select five different values for $r$: two negative, zero, and two positive. We let $t$ include the integers from 1 to 100. We then use `sapply` to apply our function of continuous exponential growth to each $r$, across all time steps. This results in a matrix where each row is the population size at each time $t$, and each column uses a different $r$.
```{r}
r <- c(-0.03, -0.02, 0, 0.02, 0.03)
N0 <- 2
t <- 1:100
cont.mat <- sapply(r, function(ri) N0 * exp(ri * t))
```

Next we create side-by-side plots, using both arithmetic and logarithmic scales, and add a legend.
```{r}
layout(matrix(1:2, nrow = 1))
matplot(t, cont.mat, type = "l", ylab = "N", col = 1)
legend("topleft", paste(rev(r)), 
       lty = 5:1, col = 1, bty = "n",
       title = "r")
matplot(t, cont.mat, type = "l", ylab = "N", log = "y", col = 1)
```

### 1.4.3 Doubling (and tripling) time

#### Creating a function for doubling time

We can create a function for this formula, and then evaluate it for different values of $m$ and $r$. For $m = 2$, we refer to this as “doubling time.” When we define the function and include arguments $r$ and $m$, we also set a default value for `m = 2`. This way, we do not always have to type a value for $m$; by default the function will return the doubling time.
```{r}
m.time <- function(r, m = 2) {
  log(m)/r
}
```

Now we create a vector of $r$, and then use `m.time` to generate a vector of doubling times.
```{r}
rs <- c(0, 1, 2)
m.time(rs)
# [1]    Inf 0.6931 0.3466
```

Note that R tells us that when $r = 0$, it takes an infinite (`Inf`) amount of time to double. This is what we get when we try to divide by zero!

## 1.6 Modeling with Data: Simulated Dynamics

### 1.6.2 Looking at and collecting the data

```{r}
names(sparrows)
# [1] "Year"           "Count"          "ObserverNumber"
attach(sparrows)
```

Now we plot these counts through time (Fig. 1.8).

```{r}
plot(Count ~ Year, type = "b")
```
```{r}
obs.R <- Count[-1]/Count[-length(Count)]
```

```{r}
plot(obs.R ~ Year[-length(Count)])
abline(h = 1, lty = 3)
```

### 1.6.3 One simulation

```{r}
years <- 50
set.seed(3)
sim.Rs <- sample(x = obs.R, size = years, replace = TRUE)

output <- numeric(years + 1)

output[1] <- Count[Year == max(Year)]

for (t in 1:years) output[t + 1] <- {
  output[t] * sim.Rs[t]
}

plot(0:years, output, type = "l")
```

### 1.6.4 Multiple simulations

```{r}
sims = 10
sim.RM <- matrix(sample(obs.R, sims * years, replace = TRUE),
                 nrow = years, ncol = sims)

output[1] <- Count[Year == max(Year)]
outmat <- sapply(1:sims, function(i) {
  for (t in 1:years) output[t + 1] <- output[t] * sim.RM[t, i]
  output
  })

matplot(0:years, outmat, type = "l", log = "y")

```

### 1.6.5 Many simulations, with a function

```{r}
PopSim <- function(Rs, N0, years = 50, sims = 10) {
  sim.RM = matrix(sample(Rs, size = sims * years, replace = TRUE),
                  nrow = years, ncol = sims)
  output <- numeric(years + 1)
  output[1] <- N0
  outmat <- sapply(1:sims, function(i) {
    for (t in 1:years) output[t + 1] <- round(output[t] * sim.RM[t, i], 0)
    output
  })
  return(outmat) 
}
```

```{r}
system.time(output <- PopSim(Rs = obs.R, N0 = 43, sims = 1000))
#  user  system elapsed
# 0.404   0.004   0.407
```

This tells me that it took less than half a second to complete 1000 simulations. That helps me understand how long 100,000 simulations might take. We also check the dimensions of the output, and they make sense.
``` {r}
dim(output)
# [1] 51 1000
```

### 1.6.6 Analyzing results

```{r}
N.2053 <- output[51, ]
summary(N.2053, digits = 6)
# Min.  1st Qu.   Median     Mean  3rd Qu.     Max.
#  0.0     14.0     66.0   1124.6    291.8 332236.0
```

``` {r}
quantile(N.2053, prob = c(0.0275, 0.975))
# 2.75% 97.5%
#     0  5967
```

```{r}
hist(N.2053, main = "N")
hist(log10(N.2053 + 1), main = "log(N+1)")
abline(v = log10(quantile(N.2053, prob = c(0.0275, 0.975)) + 1), 
       lty = 3)
```

```{r}
logOR <- log(obs.R)
n <- length(logOR)
t.quantiles <- qt(c(0.025, 0.975), df = n - 1)

se <- sqrt(var(logOR)/n)
CLs95 <- mean(logOR) + t.quantiles * se

R.limits <- exp(CLs95)
R.limits
# [1] 0.8968 1.1302

N.Final.95 <- Count[Year == max(Year)] * R.limits^50
round(N.Final.95)
# [1] 0 19528
```

```{r}
qqplot(qt(ppoints(n), df = n - 1), scale(logOR))
qqline(scale(logOR))
```