---
title: "Sean’s Chapter 5 template"
author: "Sean Denny"
date: "10/10/2019"
output:
  html_document:
    toc: false
    code_folding: show
    number_sections: false
    theme: cerulean
    highlight: haddock
  pdf_document:
    toc: false

---

```{r setup, echo = TRUE, message = FALSE, warning = FALSE}

knitr::opts_chunk$set(fig.width = 6, fig.height = 4, # fig.path = 'Figs/',
                      echo = TRUE, message = FALSE, warning = FALSE)

library(tidyverse)
library(RColorBrewer)

### The book has an R package! 
library(primer) # install.packages('primer')

```

# Chapter 5: Lotka-Volterra Interspecific Competition

## 5.1 Discrete and Continuous Time Models

### 5.1.1 Discrete time model

$$N_{t+1} = N_t + r_d N_t (1 - \alpha N_t)$$
\alpha is the per capita effect of each individual on each other, so the $$(1 - \alpha N_t)$$ portion is the density-dependence term.

Adding in interspecific competition:

$$N_{1,t+1} = N_{1,t} + r_{1,d} N_{1,t} (1 - \alpha_{11} N_{t,1} - \alpha_{12} N_{2,t})$$

in which $\alpha_{11}$ is the effect of species 1 on itself, and $\alpha_{12}$ is the effect of species 2 on the growth rate of species 1.

$N_{2_,t+1} = N_{2,t} + r_{2,d} N_{2,t} (1 - \alpha_{21} N_{1,t} - \alpha_{22} N_{2,t})$
in which $\alpha_{21}$ is the effect of species 1 on the growth rate of species 2, and $\alpha_{22}$ is the effect of species 2 on itself.

#### Code for a model of discrete logistic competition:

```{r Model of discrete logistic growth with competition}

dlv_comp <- function(N, alpha, rd = c(1,1)) #why make c (1,1) instead of just 1?
                                            {
  N1_t1 <- N[1] + rd * N[1](1 - alpha[1,1] * N[1] - alpha[1,2] * N[2])
  N2_t1 <- N[2] + rd * N[2](1 - alpha[2,2] * N[2] - alpha[2,1] * N[1])
  c(N1_t1, N2_t1)
 }  

```



  

``` {r} 
dlvcomp2 <- function(N, alpha, rd = c(1, 1)) {
  N1.t1 <- N[1] + rd[1] * N[1] * (1 - alpha[1, 1] * N[1] - 
                                    alpha[1, 2] * N[2])
  N2.t1 <- N[2] + rd[2] * N[2] * (1 - alpha[2, 1] * N[1] - 
                                    alpha[2, 2] * N[2])
  c(N1.t1, N2.t1)
}

```
Note the indices for `alpha` match the subscripts in eqs. 5.2, 5.3.







``` {r} 
dlvcomp2 <- function(N, alpha, rd = c(1, 1)) {
  N1.t1 <- N[1] + rd[1] * N[1] * (1 - alpha[1, 1] * N[1] - 
                                    alpha[1, 2] * N[2])
  N2.t1 <- N[2] + rd[2] * N[2] * (1 - alpha[2, 1] * N[1] - 
                                    alpha[2, 2] * N[2])
  c(N1.t1, N2.t1)
}

```
Note the indices for `alpha` match the subscripts in eqs. 5.2, 5.3.





}